---
title: "Stock Price Prediction"
author: "Frank Agyemang"
date: "2026-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quant Finance Series

```{r, warning=FALSE, echo=FALSE}
## Loading required libraries
library(quantmod)
#library(tidyquant)
library(tidyr)
library(dplyr)
library(ggplot2)
library(PerformanceAnalytics)
library(xts)
library(zoo)
library(mFilter)
```

## Stock Return Analysis: Fetch, Compute Returns, Plot, Decompose


```{r}
## Define tickers and date range 

tickers <- c("JPM", "KO", "XOM", "IBM")   # example NYSE stocks
start_date <- "2020-01-01"
end_date   <- Sys.Date()

##  Download price data 
getSymbols(tickers,
           src = "yahoo",
           from = start_date,
           to = end_date,
           auto.assign = TRUE)

# Combine adjusted close prices into one xts object
price_list <- lapply(tickers, function(t) Ad(get(t)))
prices <- do.call(merge, price_list)
colnames(prices) <- tickers
```




```{r}
## Calculate daily returns
# Log returns (Used here)
returns_log <- na.omit(Return.calculate(prices, method = "log"))

## Superimposed time series plot of returns 
returns_df <- returns_log %>%
  fortify.zoo() %>%          # converts xts to data frame with a Date column
  rename(Date = Index) %>%
  pivot_longer(-Date, names_to = "Ticker", values_to = "Return")

ggplot(returns_df, aes(x = Date, y = Return, color = Ticker)) +
  geom_line(alpha = 0.8, linewidth = 0.4) +
  labs(title = "Daily Log Returns",
       subtitle = paste(tickers, collapse = ", "),
       x = "Date", y = "Log Return") +
  theme_minimal() +
  theme(legend.position = "top")
```


 Daily stock returns are close to noise (efficient markets), so classical  "seasonality" is usually negligible -- this is included because it was  requested, but interpret weak seasonal/cyclical components with caution. STL (stl()) splits a series into trend + seasonal + remainder only. It does NOT separate a distinct "cyclical" component -- cycles longer than the seasonal period end up folded into the trend. To pull out a genuine business-cycle-length component, we additionally  apply an HP (Hodrick-Prescott) filter to the STL trend, which splits it into a smooth long-run trend and a cyclical deviation from it.


```{r}
##  Decomposition: trend, seasonality, cyclicality, residual  
##   This requires the `mFilter` package

decompose_stock <- function(ret_xts, ticker, frequency = 21) {
  # frequency = 21 treats a trading month as the "seasonal" period;
  # use 5 for a trading week, or 252 for an annual cycle instead.
  ts_data <- ts(as.numeric(ret_xts), frequency = frequency)

  stl_fit <- stl(ts_data, s.window = "periodic", robust = TRUE)
  comp <- as.data.frame(stl_fit$time.series)  # seasonal, trend, remainder 
  # Further split the STL trend into long-run trend + cycle via HP filter
  hp_fit <- hpfilter(comp$trend, freq = 14400)  # 14400 is a common daily-data lambda
  comp$long_run_trend <- as.numeric(hp_fit$trend)
  comp$cycle          <- as.numeric(hp_fit$cycle)

  comp$Date   <- index(ret_xts)
  comp$Ticker <- ticker
  comp
}
```



```{r}

decomp_list <- lapply(tickers, function(t) decompose_stock(returns_log [, t], t))
decomp_all  <- bind_rows(decomp_list)
```


```{r}
##  Plot each stock's decomposition 
plot_decomposition <- function(df, ticker) {
  df_long <- df %>%
    filter(Ticker == ticker) %>%
    select(Date, seasonal, long_run_trend, cycle, remainder) %>%
    rename(Trend = long_run_trend, Seasonal = seasonal,
           Cyclical = cycle, Residual = remainder) %>%
    pivot_longer(-Date, names_to = "Component", values_to = "Value")

  ggplot(df_long, aes(x = Date, y = Value)) +
    geom_line(color = "steelblue", linewidth = 0.4) +
    facet_wrap(~ Component, scales = "free_y", ncol = 1) +
    labs(title = paste("Decomposition of Daily Returns:", ticker),
         x = "Date", y = NULL) +
    theme_minimal()
}

# Example: plot decomposition for each ticker
for (t in tickers) {
  print(plot_decomposition(decomp_all, t))
}

##  (Optional) Save plots to files
ggsave("returns_superimposed.png", width = 10, height = 6)
 for (t in tickers) {
  ggsave(paste0("decomp_", t, ".png"), plot = plot_decomposition(decomp_all, t),
          width = 8, height = 8)
}
```

