---
title: "Portfolio_Optimization_Simulated_Data"
author: "Frank Agyemang"
date: "2026-06-27"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Portfolio optimization

This project simulates data to determine risk and return of a hypothetical portfolio in R.

```{r, warning=FALSE,echo=FALSE}
library(plotly)
library(dplyr)
library(quantmod)
library(ggplot2)
library(forecast)

```


```{r}
# Suppress warnings
options(warn = -1)

# Set parameters
n_assets <- 5
n_simulation <- 500

# Generate random returns matrix
returns <- matrix(
  rnorm(n_assets * n_simulation),
  nrow = n_assets,
  ncol = n_simulation
)


# Simulated returns on stocks
stock_ret <- as.data.frame(t(returns))
```


```{r}
# Finding the 5th percentile of each stock
percentile_5 <- sapply(stock_ret, quantile, probs = 0.05, na.rm = TRUE)
print(percentile_5)

# Mean return on each stock
means <- sapply(stock_ret, mean, na.rm = TRUE)
print(means)

# Variance of return on each stock
variances <- sapply(stock_ret, var, na.rm = TRUE)
print(variances)

# Covariance matrix  
cov_matrix <- cov(stock_ret, use = "complete.obs")
print(cov_matrix)

# Value at Risk assuming i have $1000
VaR_s1 <- 1000 * -1.848217
VaR_s2 <- 1000 * -1.534539
VaR_s3 <- 1000 * -1.585278
VaR_s4 <- 1000 * -1.733975
VaR_s5 <- 1000 * -1.704949
```


```{r}

# Finding the eigen value and eigen vector using the covariance matrix
eigen_decomp <- eigen(cov_matrix)
eigen_decomp$values    
eigen_decomp$vectors

# Sum eigen values
sum_eigen_vals <- sum(eigen_decomp$values)

# Proportion of variance explained by each component
prop_var <- eigen_decomp$values / sum(eigen_decomp$values)
print(prop_var)

# nb: all are greater or equal to 0
```

Random weights are generated. In portfolio optimization, the goal is to allocated limited resources in a way that maximizes returns. Weights represents the percentage or proportion of resources that is being allocated to each asset in the portfolio. In our simulation, we have a rate of return of $-0.8$ percent and a $48.5$ percent portfolio risk. That is, we stand a $48.5$ percent chance to lose about $0.8$ percent of our investment.

```{r}
## Creating a function that calculates returns for each stock 
## In the portfolio

# Generate random weights
rand <- runif(n_assets)

# Normalize weights so they add up to 1
weights <- rand / sum(rand)

port_return <- function(returns) {
  
  # Mean return for each asset
  rets <- rowMeans(returns)
  
  # Covariance matrix of asset returns
  cov_matrix <- cov(t(returns))
  
  # Portfolio return
  portfolio_returns <- sum(weights * rets)
  
  # Portfolio standard deviation
  portfolio_std_dev <- sqrt(
    t(weights) %*% cov_matrix %*% weights
  )
  
  return(list(
    portfolio_returns = portfolio_returns,
    portfolio_std_dev = portfolio_std_dev
  ))
}

# Run the function
result <- port_return(returns)

portfolio_returns <- result$portfolio_returns
portfolio_std_dev <- result$portfolio_std_dev

```

We fit a regression of standard deviation and return. The objective is to show that both risk(shown by standard deviation) and return go hand in hand. Nevertheless, the magnitude of this association depends on the assets or stocks we may have in our portfolio.

```{r}

# Create portfolio simulation for i = 1 to 100
portfolio <- t(sapply(1:100, function(i) {
  port_return(matrix(rnorm(n_assets * i), nrow = n_assets, ncol = i))
}))

# Convert to data frame
portfolio_df <- as.data.frame(portfolio)

# Renaming variables in the data frame
portfolio_df <- portfolio_df %>%
  rename(Return = portfolio_returns)
portfolio_df <- portfolio_df %>%
  rename(Standard_Deviation = portfolio_std_dev)

# Changing variable type to numeric
portfolio_df$Return <- as.numeric(portfolio_df$Return)
portfolio_df$Standard_Deviation <- as.numeric(portfolio_df$Standard_Deviation)

# Keep only rows with no missing values for the regression
valid_rows <- complete.cases(portfolio_df[, c("Return", "Standard_Deviation")])

# Fit model only on valid rows
model <- lm(Standard_Deviation ~ Return, data = portfolio_df[valid_rows, ])

# Create Best_Fit column with NA first
portfolio_df$Best_Fit <- NA_real_

# Put fitted values only in the valid rows
portfolio_df$Best_Fit[valid_rows] <- fitted(model)

# Plot risk-return relationship
fig <- plot_ly() %>%
  add_trace(
    data = portfolio_df,
    x = ~Return,
    y = ~Standard_Deviation,
    type = "scatter",
    mode = "markers",
    name = "Risk-Return Relationship"
  ) %>%
  add_trace(
    data = portfolio_df,
    x = ~Return,
    y = ~Best_Fit,
    type = "scatter",
    mode = "lines",
    name = "Best Fit Line"
  ) %>%
  layout(
    xaxis = list(title = "Return"),
    yaxis = list(title = "Standard Deviation"),
    width = 900,
    height = 470
  )

fig
```


## Time Series

```{r}
# =====================================================================
# Pull Apple (AAPL) closing price + date from Yahoo Finance and plot
# =====================================================================


#  Pull data 
getSymbols("AAPL", src = "yahoo",
           from = "2015-01-01", to = Sys.Date(),
           auto.assign = TRUE)          # creates an xts object called AAPL

#  Pull out just Date and Close as a plain data frame 
# Cl() grabs the Close column regardless of exact colname
aapl_df <- data.frame(
  Date  = index(AAPL),
  Close = as.numeric(Cl(AAPL))         
)



#  Base R version (quantmod's own chart function) 
chartSeries(AAPL, TA = NULL, name = "AAPL Closing Price")

#  ggplot2 time series plot 
ggplot(aapl_df, aes(x = Date, y = Close)) +
  geom_line(color = "steelblue") +
  labs(title = "Apple (AAPL) Closing Price", x = "Date", y = "Close ($)") +
  theme_minimal(base_size = 13)

#  Save the plot 
dir.create("images", showWarnings = FALSE)
ggsave("images/AAPL_Close.png", width = 10, height = 6, dpi = 100)
```


```{r}
## Get the components of time series

#  Pull data 
getSymbols("AAPL", src = "yahoo",
           from = "2015-01-01", to = Sys.Date(),
           auto.assign = TRUE)
AAPL_monthly <- to.monthly(AAPL, indexAt = "firstof", OHLC = FALSE)

# monthly closing price, still an xts object
aapl_close <- Cl(AAPL_monthly)     

#  Convert to a ts() object with monthly frequency (period = 12) 
aapl_ts <- ts(as.numeric(aapl_close),
            start = c(2015, 1),
            frequency = 12)

#  Decompose - additive is closest to statsmodels' default 
decomp <- decompose(aapl_ts, type = "additive")

#  Plot all four panels (observed, trend, seasonal, random) 
dir.create("images", showWarnings = FALSE)
png("images/decomposition.png", width = 1000, height = 700, res = 100)
plot(decomp)
```


```{r}
# Time series components (visualization)
# =====================================================================
# R equivalent: pull ^GSPC monthly close from Yahoo Finance and
# decompose into trend, seasonal, and remainder components
# =====================================================================


# Pull daily data, then convert to monthly (yfinance's interval='1mo')
getSymbols("^GSPC", src = "yahoo",
           from = as.Date("2015-01-01"), to = as.Date("2021-01-01"),
           auto.assign = TRUE)

GSPC_monthly <- to.monthly(GSPC, indexAt = "firstof", OHLC = FALSE)

# monthly closing price, still an xts object
sp_close <- Cl(GSPC_monthly)     

#  Convert to a ts() object with monthly frequency (period = 12) 
sp_ts <- ts(as.numeric(sp_close),
            start = c(2015, 1),
            frequency = 12)

#  Decompose - additive is closest to statsmodels' default 
decomp <- decompose(sp_ts, type = "additive")

#  Plot all four panels (observed, trend, seasonal, random) 
dir.create("images", showWarnings = FALSE)
png("images/decomposition.png", width = 1000, height = 700, res = 100)
plot(decomp)


  
```



```{r}

## Auto correlation function
p <- ggAcf(aapl_ts, lag.max = 30) +
  xlab("Number of Lags") +
  ggtitle(NULL)

ggsave("images/acf_SP.png", plot = p)
print(p)

## Partial ACF
p1 <- ggPacf(aapl_ts, lag.max = 30) +
  xlab("Number of Lags") +
  ggtitle(NULL)

ggsave("images/pacf_SP.png", plot = p)
print(p1)
```

