---
title: "Portfolio_Optimization"
author: Frank Agyemang, Dr. Blaine Johnson, Dr. Allan Fritz, Addy Carrol, Dr. Reka
  Howard
date: "2026-06-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Portfolio optimization

This project simulates data to determine risk and return of a hypothetical portfolio in R.

```{r, warning=FALSE,echo=FALSE}
library(plotly)
library(dplyr)
```


```{r}
# Suppress warnings
options(warn = -1)

# Set parameters
n_assets <- 5
n_simulation <- 500

# Generate random returns matrix
returns <- matrix(
  rnorm(n_assets * n_simulation),
  nrow = n_assets,
  ncol = n_simulation
)


```

Random weights are generated. In portfolio optimization, the goal is to allocated limited resources in a way that maximizes returns. Weights represents the percentage or proportion of resources that is being allocated to each asset in the portfolio. In our simulation, we have a rate of return of $-0.8$ percent and a $48.5$ percent portfolio risk. That is, we stand a $48.5$ percent chance to lose about $0.8$ percent of our investment.

```{r}
## Creating a function that calculates returns for each stock 
## In the portfolio

# Generate random weights
rand <- runif(n_assets)

# Normalize weights so they add up to 1
weights <- rand / sum(rand)

port_return <- function(returns) {
  
  # Mean return for each asset
  rets <- rowMeans(returns)
  
  # Covariance matrix of asset returns
  cov_matrix <- cov(t(returns))
  
  # Portfolio return
  portfolio_returns <- sum(weights * rets)
  
  # Portfolio standard deviation
  portfolio_std_dev <- sqrt(
    t(weights) %*% cov_matrix %*% weights
  )
  
  return(list(
    portfolio_returns = portfolio_returns,
    portfolio_std_dev = portfolio_std_dev
  ))
}

# Run the function
result <- port_return(returns)

portfolio_returns <- result$portfolio_returns
portfolio_std_dev <- result$portfolio_std_dev

```

We fit a regression of standard deviation and return. The objective is to show that both risk(shown by standard deviation) and return go hand in hand. Nevertheless, the magnitude of this association depends on the assets or stocks we may have in our portfolio.

```{r}

# Create portfolio simulation for i = 1 to 100
portfolio <- t(sapply(1:100, function(i) {
  port_return(matrix(rnorm(n_assets * i), nrow = n_assets, ncol = i))
}))

# Convert to data frame
portfolio_df <- as.data.frame(portfolio)

# Renaming variables in the data frame
portfolio_df <- portfolio_df %>%
  rename(Return = portfolio_returns)
portfolio_df <- portfolio_df %>%
  rename(Standard_Deviation = portfolio_std_dev)

# Changing variable type to numeric
portfolio_df$Return <- as.numeric(portfolio_df$Return)
portfolio_df$Standard_Deviation <- as.numeric(portfolio_df$Standard_Deviation)

# Keep only rows with no missing values for the regression
valid_rows <- complete.cases(portfolio_df[, c("Return", "Standard_Deviation")])

# Fit model only on valid rows
model <- lm(Standard_Deviation ~ Return, data = portfolio_df[valid_rows, ])

# Create Best_Fit column with NA first
portfolio_df$Best_Fit <- NA_real_

# Put fitted values only in the valid rows
portfolio_df$Best_Fit[valid_rows] <- fitted(model)

# Plot risk-return relationship
fig <- plot_ly() %>%
  add_trace(
    data = portfolio_df,
    x = ~Return,
    y = ~Standard_Deviation,
    type = "scatter",
    mode = "markers",
    name = "Risk-Return Relationship"
  ) %>%
  add_trace(
    data = portfolio_df,
    x = ~Return,
    y = ~Best_Fit,
    type = "scatter",
    mode = "lines",
    name = "Best Fit Line"
  ) %>%
  layout(
    xaxis = list(title = "Return"),
    yaxis = list(title = "Standard Deviation"),
    width = 900,
    height = 470
  )

fig
```



